---
title: "Analysis of the number of days an animal spends in the shelter"
author: "Group 17"
number-sections: true
format: 
  html:
    embed-resources: true
    code-tools: true
  pdf: default
editor_options: 
  chunk_output_type: console
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

# introduction

Data include animal type, month, year, intake type, outcome type, chip status, time at shelter. This study is dedicated to exploring Which factors influence the number of days an animal spends in the shelter before its final outcome is decided.

# Main body

## Import packages

```{r}
#| label: libraries
library(tidyverse)
library(moderndive)
library(gapminder)
library(sjPlot)
library(stats)
library(jtools)
library(ggplot2)
library(MASS)
library(car)
library(pscl)

```

## Read data and plot

```{r}
# read data
data <- read.csv("dataset17.csv", header = TRUE)
head(data)

animal<- data %>%
  filter(animal_type == "DOG" |  animal_type=="CAT")%>%
  droplevels()

animal <- animal%>%
  mutate(animal_type=as.factor(animal_type),
         intake_type=as.factor(intake_type),
         outcome_type=as.factor(outcome_type),
         chip_status=factor(chip_status,levels=c("UNABLE TO SCAN","SCAN CHIP","SCAN NO CHIP")),
         month =as.factor(month),
         year=as.factor(year))

# check the factors
levels(animal$animal_type)
levels(animal$intake_type)
levels(animal$outcome_type)
levels(animal$chip_status)

```

```{r}
#plot histogram
#| fig-cap: Histogram of the number of days an animal spends in the shelter
#| label: fig-1
#| fig-width:  8
#| fig-height: 6
#| fig-align: center
ggplot(animal, aes(x = time_at_shelter)) +
  geom_histogram(binwidth = 2, fill = "skyblue", color = "black", alpha = 0.8) +
  labs( x = "time_at_shelter", y = "Number") +
  theme_minimal()

sum(animal$time_at_shelter == 0)
mean(animal$time_at_shelter == 0)

```

The @fig-1 presents a right-skewed distribution. Most animals stay in shelters for a relatively short period of time, concentrated   in the range of 0-10 days, and very few animals stay for more than 30 days or even longer.And 374 observations (22.58%) have a `time_at_shelter` value of 0, suggesting that it should be carefully considered when fitting the model.

```{r}
# boxplot
#| fig-cap: Boxplot of time_at_shelter and animal_type
#| label: fig-2
#| fig-width:  8
#| fig-height: 6
#| fig-align: center
ggplot(data = animal, aes(x = animal_type, y = time_at_shelter, fill = animal_type)) +
  geom_boxplot() +
  labs(x = "animal type", y = "time at shelter")+ 
  theme(legend.position = "none")
```

The median of cats is smaller than that of dogs, but dogs have a wider distribution. The upper quartile range for dogs is larger, meaning that some dogs stay in shelters for longer periods of time. 

```{r}
#| fig-cap: Boxplot of time_at_shelter and intake_type
#| label: fig-3
#| fig-width:  8
#| fig-height: 6
#| fig-align: center
#boxplot_2
ggplot(animal, aes(x = intake_type, y = time_at_shelter,fill = intake_type)) +
  geom_boxplot() +
  theme_minimal()
```

The median of confiscated animals was significantly higher than that of the other two groups, indicating that confiscated animals stayed concentrated and longer.

```{r}
#| fig-cap: Boxplot of time_at_shelter and chip_status
#| label: fig-4
#| fig-width:  8
#| fig-height: 6
#| fig-align: center
#boxplot_3
ggplot(animal, aes(x = chip_status, y = time_at_shelter,fill = chip_status)) +
  geom_boxplot() +
  theme_minimal()
```

The medians of the three sets of data were relatively close, indicating that animals with different chip statuses stayed in the shelter for roughly similar lengths of time. The IQR of the animals scanned to the chip was wider, indicating that the data distribution in this group was more dispersed.

```{r}
# Scatter Plot
#| fig-cap: Boxplot of time_at_shelter and intake_type
#| label: fig-5
#| fig-width:  8
#| fig-height: 6
#| fig-align: center
ggplot(animal, aes(x = factor(month, levels = c(10:12, 1:9)), y = time_at_shelter, color = factor(year))) +
  geom_jitter(alpha = 0.6, width = 0.2) +  
  labs(title = "time_at_shelte vs. Month", x = "Month", y = "time_at_shelte", color = "Year") +
  theme_minimal()
```

The time range of the dataset is from October to December 2016 and January to September 2017. The monthly data is complete so we consider it as an explanatory variable. The length of time animals spent in shelters did not show significant seasonal variations across years, with most animals staying short (less than 20 days).

## model and anaylsis

### Poisson regression

Poisson regression is one of the most commonly used generalized linear models when analyzing numerical data, so the study starts with poisson regression.

```{r}

# set up "month" in order
animal <- animal %>%
  mutate(month = as.numeric(month),
         month_ordered = ifelse(month >= 10, month - 9, month + 3))


# Poisson regression
glm_model <- glm(time_at_shelter ~ animal_type +  month_ordered + intake_type  + chip_status,
                 data = animal,
                 family = poisson(link = "log"))

summary(glm_model)

# overdivergence of Poisson regression
dispersion <- sum(residuals(glm_model, type = "pearson")^2) / glm_model$df.residual
dispersion

```

The dispersion parameter (8.007) is much larger than 1, which means our GLM may have overdispersion. Therefore, we consider negative binomial regression to improve the model's fit.


### Negative binomial regression model

Negative binomial distribution is used when the data is too discrete

```{r}
# Negative binomial regression model 1
nb_model_1 <- glm.nb(time_at_shelter ~ animal_type + intake_type  +chip_status + month_ordered ,data = animal)
summary(nb_model_1)

# Fit the full model (including all explanatory variables)
full_model <- glm.nb(time_at_shelter ~ animal_type + intake_type  +chip_status + month_ordered , data = animal)

# Fit an empty model (only intercept terms)
null_model <- glm.nb(time_at_shelter ~ 1, data = animal)

# Model selection using stepwise regression (two-way selection), based on AIC index
selected_model <- stepAIC(null_model, 
                          scope = list(lower = null_model, upper = full_model), 
                          direction = "both")

summary(selected_model)

# Negative binomial regression model 2
nb_model_2 <- glm.nb(time_at_shelter ~  intake_type  +chip_status + month_ordered ,data = animal)

summary(nb_model_2)

```




### **Zero-inflated model**

Considering that the sample contains too many zeros, we decided to try Zero-inflated model.

```{r}
# Zero-inflated model

zeroflated_model <- zeroinfl(time_at_shelter ~ animal_type + intake_type + chip_status + month_ordered, data = animal, dist = "negbin")

summary(zeroflated_model)

# AIC, BIC and mcfadden R²

aic_value <- AIC(zeroflated_model)
bic_value <- BIC(zeroflated_model)
print(aic_value)
print(bic_value)


null_model <- update(zeroflated_model, . ~ 1)
mcfadden_r2 <- 1 - as.numeric(logLik(zeroflated_model)) / as.numeric(logLik(null_model))
print(mcfadden_r2)

# Comment:
# AIC is 9335 and BIC is 9416 , both of them are lower than the model above. Mcfadden R² is 0.013

```

###Model comparison

```{r}
# AIC
aic_values <- AIC(glm_model,nb_model_1, nb_model_2)

# Deviance
deviance_values <- c(deviance(glm_model), deviance(nb_model_1), deviance(nb_model_2))

#df
df_values <- c(glm_model$df.residual, nb_model_1$df.residual, nb_model_2$df.residual)

# merge
model_comparison <- data.frame(
  Model = c("GLM (Poisson)", "NB1 (Negative Binomial)", "NB2 (Negative Binomial)"),
  Deviance = deviance_values,
  DF = df_values,
  AIC = aic_values$AIC
)

model_comparison
```

By comparing AIC, the negative binomial regression model performs better.

## Residual Plot

```{r}
animal$residuals <- residuals(glm_model, type = "deviance")
predicted_values <- predict(glm_model, type = "response")

ggplot(animal, aes(x = predicted_values, y = residuals)) +
  geom_point(alpha = 0.4, color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs. Predicted Values", x = "Predicted Values", y = "Residuals") +
  theme_minimal()
```

The variability of the residuals is larger at smaller predicted values, which indicates excessive dispersion, i.e., the variance is larger than the mean of Poisson's assumptions. This is consistent with the dispersion parameter we calculated earlier.

### nb_model_2

```{r}
animal$residuals <- residuals(nb_model_2, type = "deviance")
predicted_values <- predict(nb_model_2, type = "response")

ggplot(animal, aes(x = predicted_values, y = residuals)) +
  geom_point(alpha = 0.4, color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs. Predicted Values", x = "Predicted Values", y = "Residuals") +
  theme_minimal()

```

The variance of the residuals is generally in the range of -2 to +2. The negative binomial regression model basically solves the overdispersion problem.

### zeroflated_model

```{r}
res <- residuals(zeroflated_model, type = "pearson")
fitted_vals <- fitted(zeroflated_model)

ggplot(animal, aes(x = fitted_vals, y = res)) +
  geom_point(alpha = 0.4, color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs. Predicted Values", x = "Predicted Values", y = "Residuals") +
  theme_minimal()

```

# summary

The report provides an in-depth statistical analysis of the number of days animals spend in shelters and uses a variety of models to reveal the patterns behind the data. Overall, the data show a clear right-skewed distribution: most animals have shorter stays (mainly between 0 and 10 days), while a few animals may have longer stays.

In order to explore the factors affecting the length of stay of animals, Poisson regression and negative binomial regression models were first used to analyze the role of variables such as animal entry method, exit result, animal species, chip state and month order. The results show that "OWNER SURRENDER" and "STRAY" are often associated with shorter stay days; In the results, "DIED", "EUTHANIZED" and "RETURNED TO OWNER" also significantly correspond to a shorter residence time, while dogs tend to stay longer, although chip status and month order have an impact, but the effect is weak.

Due to the large number of zeros in the data, the report further introduces the zero expansion model, and the negative binomial model is better on both AIC and BIC.

# Furture work

Considering the large number of 0 values, the presence of multiple explanatory variables in the data, and the fact that glm's fitting is not very good, we can try to use ANN to predict the time of animals in shelters.

How to transform non-numerical explanatory variables into numerical explanatory variables is still in the process of research, or will the frequency of observer occurrence be considered as the conversion criterion.

